---
output:
  pdf_document:
    includes:
      in_header: preamble.tex

  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

source("1-load-data.R")
source("2-data-preparation.R")
source("3-modeling.R")

library(RColorBrewer)
library(sjPlot)
```

# Policy brief: Critical Thinking, Media, and Democracy in an Age of Unchecked Technological Development

\textit{Document created `r Sys.Date()`}

## Education against Technological Intrusion

The ever-accelerating digital technology development brings convenience and entertainment, but creates new avenues for exploitation and manipulation. It may threaten democracy through the spread of fake or misleading information on the Internet. As digital technology is part of everyday life, it it needs to be addressed through education so that people can use it competently and understand its limitations and impact on society.

\begin{center}
\begin{tcolorbox}[orangebox]
This brief seeks to answer the following questions:

\begin{enumerate}
  \item What influences Europeans' view that \textbf{critical thinking, media, and democracy should be taught more in compulsory education} in a time of rapid technological development?
  \item What \textbf{policy recommendations can be formulated} to encourage this view?
\end{enumerate}
\end{tcolorbox}
\end{center}

## The Demand for Civic Education

\begin{flushleft}
In the CRONOS3 survey, Europeans from 11 countries were asked \textit{"Which topics would you like to see being taught more in compulsory education?"}. They could choose up to 3 answers from a list of 6 categories, one of which was \textbf{"Critical thinking, media and democracy"}. \textcolor{exactgreen}{\textbf{Across participating countries, as many as 73\% of Austrians}} and \textcolor{exactorange}{\textbf{as few as 31\% of Hungarians}} selected this option. We analyzed 21 factors related to STEM education, modern technology use, and demography to understand what drives this support.
\end{flushleft}

## Factors of Support for Critical Thinking, Media and Democracy Education

```{r logit-plot, fig.height=3.5, message=FALSE}

non_sig_MFIN <- c(
  "factor(CTR)Belgium", "factor(CTR)Czechia", "factor(CTR)Finland",
  "factor(CTR)France", "factor(CTR)United Kingdom", "factor(CTR)Hungary",
  "factor(CTR)Iceland", "factor(CTR)Poland", "factor(CTR)Portugal",
  "factor(CTR)Slovenia",
  "GND", "AGE", "HIN", "UNA", "SML", "SMP", "SMM", "SMU", "SMF",
  "AUD", "ADA", "AIR"
)

lbl_MFIN <- c(
  EDY = "Years of education",
  SMS = "Scrolls soc. media\nat least daily",
  SMR = "Reacts to soc. media\nat least daily",
  DTP = "Taught more: Digital\ntools & programming",
  MAS = "Taught more: Maths\nand sciences",
  CTA = "Negative view of\ntechnology's future",
  STR = "Trust in scientists",
  GTR = "General trust\nin others"
)

# p1 <- plot_model(
#   MFIN,
#   show.values = T,
#   value.offset = 0.4,
#   colors = "Set2",
#   rm.terms = non_sig_MFIN,
#   sort.est = T,
#   axis.labels = "",
#   axis.lim = c(0.3, 2.2),
#   title = ""
# ) +
#   labs(
#     subtitle = "Negative view of technology's future and reacting to social media\nare the strongest positive drivers; preferring more STEM education means\n lower support for critical thinking, media and democracy education.",
#     caption = "Source: Own elaboration based on CRONOS3 and ESS10-11 survey data.\nDots represent odds ratio values; lines indicate the uncertainty of the results."
#   ) +
#   theme_minimal() +
#   geom_hline(yintercept = 1, linetype = "dashed", color = "darkgray", alpha = 0.6) +
#   annotate("text", x = 3, y = 0.3,
#            label = "Negative effect", color = "#66C2A5", alpha = 0.6) +
#   annotate("text", x = 6, y = 3,
#            label = "Positive effect", color = "#FC8D62", alpha = 0.6) +
#   scale_x_discrete(labels = lbl_MFIN)
# 
# plot(p1)

# extract coefficients from model
plot_data <- data.frame(
  term = names(coef(MFIN)),
  or = exp(coef(MFIN)),
  lower = exp(confint(MFIN)[, 1]),
  upper = exp(confint(MFIN)[, 2])
) %>%
  filter(term %in% names(lbl_MFIN)) %>%
  mutate(
    pct = (or - 1) * 100,
    lower_pct = (lower - 1) * 100,
    upper_pct = (upper - 1) * 100,
    label = lbl_MFIN[term],
    label = factor(label, levels = label[order(pct)])
  )

p2 <- ggplot(plot_data, aes(x = label, y = pct, color = pct > 0)) +
  geom_point(size = 6) +
  geom_errorbar(aes(ymin = lower_pct, ymax = upper_pct), width = 0.2) +
  geom_label(
  aes(label = round(pct)),
  hjust = 0.5,
  vjust = 0.5,
  size = 2) +
  annotate("text", x = 4, y = -32,
           label = "Negative effects", color = "#66C2A5", alpha = 0.6) +
  annotate("text", x = 6, y = 71,
           label = "Positive effects", color = "#FC8D62", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray", alpha = 0.6) +
  coord_flip() +
  scale_color_manual(values = c("#66C2A5", "#FC8D62"), guide = "none") +
  labs(
    y = "Change in likelihood (%)",
    x = "",
    subtitle = "Negative view of technology's future and reacting to social media\nare the strongest positive drivers; preferring more STEM education means\n lower support for critical thinking, media and democracy education.",
    caption = "Source: Own elaboration based on CRONOS3 and ESS10-11 survey data.\nDots represent percentage change in likelihood; lines indicate uncertainty."
  ) +
  theme_minimal()

plot(p2)
```

\small This work is licensed under CC BY 4.0. To view a copy of this license, visit [creativecommons.org/licenses/by/4.0](https://creativecommons.org/licenses/by/4.0/)

\begin{center}
\includegraphics[width=24mm]{by.png}
\end{center}
```{r supl-plot, message=FALSE}
library(plotly)

lbl_full <- c(
  EDY = "Years of education completed",
  SMS = "Scrolls social media at least daily",
  SMR = "Reacts to social media at least daily",
  DTP = "Wants more: Digital tools & programming",
  MAS = "Wants more: Maths and sciences",
  CTA = "Negative view of technology's future",
  STR = "Trust in scientists",
  GTR = "General trust in others",
  GND = "Gender (female)",
  AGE = "Age",
  HIN = "Household income decile",
  UNA = "Uses internet almost all the time",
  SML = "Limits access to own social media",
  SMP = "Posts on social media at least weekly",
  SMM = "Social media exposes to misinformation",
  SMU = "Social media undermines privacy",
  SMF = "Social media prevents work/family duties",
  AUD = "Intentional digital avoidance",
  ADA = "Adapts well to technological advancements",
  AIR = "Can recognize AI-generated content",
  `factor(CTR)Belgium` = "Country: Belgium",
  `factor(CTR)Czechia` = "Country: Czechia",
  `factor(CTR)Finland` = "Country: Finland",
  `factor(CTR)France` = "Country: France",
  `factor(CTR)United Kingdom` = "Country: United Kingdom",
  `factor(CTR)Hungary` = "Country: Hungary",
  `factor(CTR)Iceland` = "Country: Iceland",
  `factor(CTR)Poland` = "Country: Poland",
  `factor(CTR)Portugal` = "Country: Portugal",
  `factor(CTR)Slovenia` = "Country: Slovenia"
)

brief_predictors <- c("EDY", "SMS", "SMR", "DTP", "MAS", "CTA", "STR", "GTR")

plot_data_full <- data.frame(
  term = names(coef(MFIN)),
  or = exp(coef(MFIN)),
  lower = exp(confint(MFIN)[, 1]),
  upper = exp(confint(MFIN)[, 2])
) %>%
  filter(term %in% names(lbl_full)) %>%
  mutate(
    pct = (or - 1) * 100,
    lower_pct = (lower - 1) * 100,
    upper_pct = (upper - 1) * 100,
    label = lbl_full[term],
    significant = !(lower_pct < 0 & upper_pct > 0),
    group = case_when(
      grepl("Country", lbl_full[term]) ~ "Country (ref: Austria)",
      term %in% brief_predictors ~ "Discussed in brief",
      TRUE ~ "Tested, not discussed"
    ),
    label = factor(label, levels = label[order(pct)]), 
    point_color = if_else(pct > 0, "#FC8D62", "#66C2A5")
  )

shape_map <- c(
  "Discussed in brief" = 16,       # circle
  "Tested, not discussed" = 15,    # square
  "Country (ref: Austria)" = 17    # triangle
)


p3_gg <- ggplot(plot_data_full,
                aes(x = label, y = pct,
                    shape = group,
                    text = paste0(
                      "<b>", label, "</b><br>",
                      "Change in likelihood: ", round(pct), "%<br>",
                      "95% CI: [", round(lower_pct), "%, ", round(upper_pct), "%]<br>",
                      if_else(significant, "Statistically significant", "Not significant")
                    ))) +
  geom_point(size = 3, color = plot_data_full$point_color) +
  geom_errorbar(aes(ymin = lower_pct, ymax = upper_pct), width = 0.3, color = plot_data_full$point_color) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray", alpha = 0.6) +
  annotate("text", x = 4, y = -32,
           label = "Negative effects", color = "#66C2A5", alpha = 0.6) +
  annotate("text", x = 26, y = 71,
           label = "Positive effects", color = "#FC8D62", alpha = 0.6) +
  coord_flip() +
#  guides(shape = guide_legend(override.aes = list(color = "gray40"))) +
  scale_shape_manual(values = shape_map, name = "") +
  labs(
    title = "Support for critical thinking, media and democracy education: all predictors",
    subtitle = "Negative view of technology's future and reacting to social media\nare the strongest positive drivers; preferring more STEM education means\nlower support for critical thinking, media and democracy education.",
    y = "Change in likelihood (%)",
    x = "",
    caption = paste0(
      "Source: Own elaboration based on CRONOS3 and ESS10-11 survey data.\n",
      "Dots represent percentage change in likelihood; lines indicate uncertainty.\n",
      "CC BY 4.0 \u2014 creativecommons.org/licenses/by/4.0\n",
      "Code and data: github.com/jakkada/I4NG-Datathon-2026"
    )
  ) +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0, size = 7),
    plot.subtitle = element_text(size = 9)
  )

p3 <- ggplotly(p3_gg, tooltip = "text") %>%
  layout(
    hoverlabel = list(bgcolor = "white"),
    margin = list(l = 220, b = 100),
    legend = list(orientation = "v", x = 1.02, y = 1, xanchor = "left")
  )

p3 <- p3 %>%
  layout(
    margin = list(l = 220, b = 140, r = 160),
    annotations = list(
      list(
        text = "Source: CRONOS3 & ESS10-11 | CC BY 4.0 | <a href='https://github.com/jakkada/I4NG-Datathon-2026'>github.com/jakkada/I4NG-Datathon-2026</a>",
        xref = "paper", yref = "paper",
        x = 0, y = -0.15,
        showarrow = FALSE,
        font = list(size = 9, color = "gray40"),
        align = "left",
        xanchor = "left"
      )
    )
  )


# save as standalone html
htmlwidgets::saveWidget(p3, "supl_p3_interactive.html", selfcontained = TRUE)
```

## Value of Normative Orientations

The most impactful factor is education: \textbf{with every completed year, the likelihood of supporting more critical thinking, media, and democracy education increases by 7\%}. A person who finishes high school is over 80\% more likely to see the value of such education than someone with no formal education.

Next are trust in scientists and general trust in others, which increase the likelihood of valuing this type of education by \textbf{13\% and 11\% per each step up}. Most striking is the effect of views about the changes technology will bring to society. \textbf{Those who expect them to be negative are 67\% more likely to support an increase in critical thinking, media and democracy education}.

\begin{center}
\begin{tcolorbox}[greenbox]
Recommendation 1:

\textbf{Involve authorities from expert and scientific communities in debates on technology and its impact on everyday life, including both positive and negative examples}
\end{tcolorbox}
\end{center}

## STEM Education Does Not Go Along with Critical Thinking, Media and Democracy

The most surprising results indicate that respondents who want more \textbf{"Digital tools and programming"} and \textbf{"Maths and sciences"} in compulsory education are also respectively \textbf{49\% and 35\% less likely} to support increased critical thinking, media and democracy education. Respondents may perceive STEM and civic-oriented education as competing paths.

\begin{center}
\begin{tcolorbox}[orangebox]
Recommendation 2:

\textbf{Introduce in STEM education curricula more critical thinking and civic-oriented training to raise awareness of social responsibility related to technology development. Introduce technology-focused elements in civic-education curricula to increase understanding of its impacts and limitations.}
\end{tcolorbox}
\end{center}

## Engagement with Social Media Matters

Scrolling through social media feeds at least once a day \textbf{decreases by 25\%} support for increased critical thinking, media and democracy education. By contrast, reacting to social media content (e.g. commenting) at least once a day \textbf{increases it by 44\%}. Contributing to social media raises awareness of the need for civic competencies, while passive consumption of its content decreases it.

\begin{center}
\begin{tcolorbox}[greenbox]
Recommendation 3:

\textbf{Access to social media should be limited or controlled among the youngest and relaxed over time, following training in how to safely and responsibly engage with it.}
\end{tcolorbox}
\end{center}

## The Importance of Country Context

All previous results hold regardless of country context. Nevertheless, countries differ significantly in their support for critical thinking, media and democracy education. \textbf{In Hungary the likelihood of support is 73\% lower, in Czechia 72\% lower, and in Poland and Portugal 71\% lower than in Austria, the reference country.} This indicates the importance of national context.

### Analytical note

\begin{flushleft}
\footnotesize

We studied 21 factors of support for critical thinking, media and democracy education. Factors with no meaningful impact included: gender, age, household income, internet use frequency, posting on social media, social media attitudes, intentional digital avoidance, and ability to recognize AI-generated content.

Open data from CRONOS3 Wave 1 (\url{https://doi.org/10.21338/cronos3-w1}), ESS10 (\url{https://doi.org/10.21338/ess10e03_2}), ESS10sc (\url{https://doi.org/10.21338/ess10sce03_1}), and ESS11 (\url{https://doi.org/10.21338/ess11e03_0}) represent countries: Austria, Belgium, Czechia, Finland, France, Hungary, Iceland, Poland, Portugal, Slovenia, and the United Kingdom. A total unweighted sample of \(n = 4{,}906\) was used in the analysis.

GNU R code to reproduce our analysis are openly available at: \url{https://github.com/jakkada/I4NG-Datathon-2026}
\normalsize
\end{flushleft}

**Authors**: Jakub Adamski, [jakub.adamski\@uni.lodz.pl](mailto:jakub.adamski@uni.lodz.pl) & Remigiusz Żulicki, [remigiusz.zulicki\@uni.lodz.pl](mailto:remigiusz.zulicki@uni.lodz.pl)
from [Faculty of Economics and Sociology](https://www.eksoc.uni.lodz.pl/en/about-the-faculty/get-to-know-eksoc), University of Lodz, Poland